Depression and suicidality as evolved credible signals of need in social conflicts

### 2021-06-14 Paper.rmd
Minor edits

In contrast to these views, we will argue that the depressed have suffered genuinely severe adversities and are therefore genuinely in need, but often have conflicts with their social partners. In these circumstances, costly and putatively dysfunctional depressive behaviors can instead be understood as aversive but credible and adaptive signals of need that elicit more support than verbal requests, sad expressions, and crying when there are conflicts of interest.

@@

Contrary to the view that the depressed have a distorted perception of their environment, there is strong evidence that most cases of depression are caused by genuinely severe negative life events, such as physical assault and death of a loved one [@Hammen2005; @Mazure1998; @devries2011; @ellsberg2008; @devries2013]. Compared to non-depressed individuals, those with depression report about twice as many instances of negative events than non-depressed individuals [@Mazure1998] and more negative events than those with schizophrenia and bipolar depression across multiple studies [@Paykel1994]. Longitudinal studies indicate that depression onsets soon after a negative event [@lewinsohn1988; @KendlerEtAl1999; @rich2005; @han2019; @sen2010] or coincides with periods where adversity is likely to increase prior to it [e.g., depression starting before a divorce rather than after, @blekesaune2008; @metsa2013; @rosenstrom2017]. Although the relationship between negative events and depression is likely bidirectional [i.e., depression probably also causes adversity, @wichers2012], negative events predict depression even when considering only events outside of one’s control, indicating that the connection is unlikely to be driven solely by individuals who are already depressed selecting into situations where negative events are likely to be common [@Hammen2005; @KendlerEtAl1999]. Furthermore, twin studies have shown that one's history of negative events remains a strong predictor of major depression when controlling for genetic similarity, and that part of the heritability of depression stems from the heritability of negative events like divorce and family conflict [@KendlerBaker2007; @KendlerEtAl1999].

@@

The missing piece of the puzzle is that depression is closely associated with anger and conflict [@cassiello2016]. Of the adversity risk factors for depression, those that involve conflict tend to be the strongest [@Hammen2005; @Mazure1998]. Marital problems, bullying, and abusive relationships are all common risk factors for depression [@KendlerEtAl1995; @KendlerEtAl1999; @klomek2007; @klomek2019], with sexual and non-sexual assault, in particular, greatly increasing one's risk of depression [@KendlerEtAl1995; @KendlerEtAl1999]. This holds true even in a small-scale, non-Western society: among the Tsimane, Amazonian horticulturalists, depression is also associated with conflict, especially conflict involving non-kin [@StieglitzEtAl2015a]. See @SymeHagen2021 for a review of the association of depression with anger and conflict.

Other notable depression risk factors, like loss of a loved one or severe or prolonged illness, might seem less related to conflict. In these situations, however, the fitness costs that stem from reduced access to resources could be mitigated with help from social partners [@SugiyamaSugiyama2003]. However, social partners might not be able, or want, to provide more investment than they already are. Indeed, there is increasing evidence that loss of a loved one is often followed by increased family conflict [see @SymeHagen2021 for a brief review]. Problems whose solutions require substantial investment or other changes on the part of social partners will therefore often involve social conflict even if they did not start that way [@Hagen2003].

Severe adversity, which often requires substantial help from social partners that they might be reluctant to provide, combined with social conflict, often of the he-said-she-said variety that could undermine victims' credibility, render "cheap" signals of need, such as sad expressions and crying, ineffective at increasing support. We argue next that some of the most costly and mysterious symptoms of depression -- profound loss of interest in virtually all activities and suicidality -- are *credible* and adaptive signals of need when there are conflicts of interest between victims of adversity and their social partners.

@@

With a cooperative species like our own, ubiquitous conflicts of interests means that there will always be disagreement over the levels of investment in a cooperative endeavor and the division of the resulting benefits. According to partner choice models, individuals who are dissatisfied with the terms of cooperation can switch partners [@hammerstein2016], e.g., workers unhappy with their pay can look for a better job. In many cases, however, it is difficult or impossible to switch partners. Spouses who are dissatisfied with their partner's investment in their new infant, for example, cannot easily find a different partner to invest in that infant [@Hagen:1999]. Similarly, an adolescent who is dissatisfied with her parent's investment in her cannot easily find other parents who were willing to invest more, nor could parents easily produce another adolescent. In these latter examples, and many cooperative endeavors central to human biological fitness, all parties have monopoly power over the benefits they bring to the endeavor -- no one is easily replaced [@Hagen2002; @Hagen2003]. Such interdependence is increasingly recognized to be important to the evolution of cooperation in humans and other animals [@roberts2005; @tomasello2012; @aktipis2018; @balliet2017].

@Hagen2003 proposed that physical aggression and core depression symptoms like loss of interest in virtually all activities were complementary strategies to resolve conflicts in interdependent relationships. @SellEtAl2009 found that physically formidable individuals were more prone to anger, prevailed more in conflicts of interest, and considered themselves entitled to better treatment. Physically or socially weaker individuals, though, are not without options to resolve conflicts in their favor. An individual with monopoly power over the benefits she contributes to a critical cooperative endeavor can withhold those benefits, or put them at risk, until her partners change their behaviors in ways that benefit her. Depression is characterized, in part by a profound loss of interest in virtually all activities [@dsm5], a symptom that is probably responsible for a sharp reduction in productivity [for review, see @SymeHagen2021]. Depression might therefore be an evolved bargaining strategy for relatively powerless individuals in the wake of adversity and social conflict [@Hagen2003; @SymeHagen2021; see also @Watson:2002].

Bargaining models assume that delaying cooperation is costly so that there is an incentive to quickly agree on a division of benefits, and more so for those who highly value the fruits of the cooperative endeavor. In a classic non-cooperative game theory model of bargaining, @rubinstein1982 showed that two parties can come to an immediate agreement over division of benefits despite conflicts of interest if the parties' valuations of cooperation are not private information: in this case, each party knows exactly what division of benefits the other will accept, and can therefore make that offer immediately, avoiding the cost of delay.

If valuations are private information, however, then costly delays might be unavoidable because each party has an incentive to request more than their actual valuation, and to reject the likely inflated requests from partners, leading to multiple rounds of bargaining. Models of bargaining with private information have a close relationship to models of credible signaling. A willingness to delay (i.e., refuse offers) credibly reveals one's low valuation of the endeavor, and therefore genuine need -- the benefit of waiting for a better offer outweighs the low cost of delay. Eagerness to reach a deal, on the other hand, credibly reveals a high valuation -- the benefit of waiting for a better offer does not outweigh the higher cost of delay. Once valuations are known, the game reduces to the one in @rubinstein1982, and the parties can reach an agreement on the division of benefits [@KennanWilson1993]. Delays also typically require that additional factors come into play [@feinberg2005 and references therein].

@@

Suicidal ideation is one of nine diagnostic criteria for a major depressive episode (MDE) [@dsm5] and is associated with depression across cultures [@haroz2017]; depression is a major risk factor for suicidal behavior [@hawton2013]; and suicidality has been invoked to justify grief-related depression symptoms as psychopathological [e.g., @pies2014]. Theoretical models of depression must account for suicidality.

Anthropology has long viewed suicidality as largely the result of social problems. Early in the field's history, anthropologists reported on suicide attempts and deaths in the small-scale societies that serve as models for the types of societies in which humans evolved. Suicide, they found, was commonly a form of protest, revenge, and/or appeal [@firth1936; @Firth:1961; @Malinowski:1932; @niehaus2012]. Some ethnographers emphasized suicide as a form of anger or social pressure [@Hezel:1987; @Giddens:1964], whereas others emphasized the powerlessness of suicide victims [@Counts:1980].

@@

<!-- Need to integrate this -->
As with depression, there are negative social responses to suicidal behavior. Most studies examining responses to those who have survived suicide attempts have focused on stigmatization. In these studies, perceptions of survivors as being weak, selfish, mentally ill, and antisocial are commonly reported [@BatterhamEtAl2013; @TzengLipson2004], with stigmatization often being found within and outside one's social network [@FreyEtAl2016; @ScoccoEtAl2012].

@@

In summary, we propose that depression symptoms that put one's value to others at risk, such as loss of interest and suicidality, are aversive but adaptive signals of need [for similar views, see @andrews2006; @HagenEtAl2008; @nock2008; @Stengel1956; @Rosenthal1993; @firth1936; @Firth:1961; @farberow1961].

<!-- Compared to suicide, however, there are fewer anthropological investigations of depression [@kleinman1985] despite frequent reports of depression or depression-like behavior in a wide range of societies [@haroz2017; @liu2020] and the potential for cross-cultural research to help understand depressive states [@haroz2017]. -->

<!-- Durkheim's pioneering study of suicide [@durkheim1897] helped launch sociology as a quantitative discipline. Sociological investigations of suicide flourished through much of the 20th century, but were supplanted by medical and public health approaches as the biomedical model of mental health took hold in the late 1970's and early 1980's [@wray2011; @syme2020]. The biomedical model tends to frame suicide as a problem with the individual, e.g., as one involving psychiatric disturbance, drug addiction, or stress sensitivity [@wray2011]. This theoretical approach has made little progress: rates of suicide, which vary widely by country, have remained relatively flat in countries for which there are long-term data, such as the US, UK, Japan, and Sweden [@owidsuicide]. -->

<!-- Evolutionary theories of depression tend to either ignore suicidality, or take it as evidence that depression is dysfunctional. Wakefield, for instance, who has devoted much of his career to arguing that most cases of depression are simply severe but normal sadness and grief, nevertheless takes suicidality to indicate pathology for some cases of depression [e.g., @wakefield2017]. @nettle2004 argues that the costs of depression, including suicidality, outweigh any benefits, and therefore adaptationist models of depression are implausible. @gilbert2006 sees suicidality as an example of a dysregulated escape defense [for reviews of evolutionary theories of depression, see @Hagen2011]. -->

<!-- As with suicide, research on depression since the 1980's has been dominated by biomedical approaches of limited success [@patten2016; @ormel2019].  -->

@@

<!-- A secondary aim was to investigate a scenario that is common in traditional societies but uncommon in Western societies -- arranged marriages -- where conflicts between parents and offspring often result in depression and suicidality [@SymeEtAl2016] -->

>>>>

In contrast to these views, we will argue that the depressed have suffered genuinely severe forms of adversity and are therefore genuinely in need, but often have conflicts with their social partners. In these circumstances, costly and putatively dysfunctional depressive behaviors can instead be understood as aversive but credible and adaptive signals of need that elicit more support than verbal requests, sad expressions, and crying when there are conflicts of interest.

@@

Contrary to the view that the depressed have a distorted perception of their environment, there is strong evidence that most cases of depression are caused by genuinely severe negative life events, such as physical assault and death of a loved one [@Hammen2005; @Mazure1998; @devries2011; @ellsberg2008; @devries2013]. Compared to non-depressed individuals, those with depression report about twice as many negative events [@Mazure1998] and more negative events than those with schizophrenia and bipolar depression across multiple studies [@Paykel1994]. Longitudinal studies indicate that depression onsets soon after a negative event [@lewinsohn1988; @KendlerEtAl1999; @rich2005; @han2019; @sen2010] or coincides with periods where adversity is likely to increase prior to it [e.g., depression starting before a divorce rather than after, @blekesaune2008; @metsa2013; @rosenstrom2017]. Although the relationship between negative events and depression is likely bidirectional [i.e., depression probably also causes adversity, @wichers2012], negative events predict depression even when considering only events outside of one’s control, indicating that the connection is unlikely to be driven solely by individuals who are already depressed selecting into situations where negative events are likely to be common [@Hammen2005; @KendlerEtAl1999]. Furthermore, twin studies have shown that one's history of negative events remains a strong predictor of major depression when controlling for genetic similarity, and that part of the heritability of depression stems from the heritability of negative events like divorce and family conflict [@KendlerBaker2007; @KendlerEtAl1999].

@@

The missing piece of the puzzle is that depression is closely associated with anger and conflict [@cassiello2016]. Of the adversity-related risk factors for depression, those that involve conflict tend to be the strongest [@Hammen2005; @Mazure1998]. Marital problems, bullying, and abusive relationships are all common risk factors for depression [@KendlerEtAl1995; @KendlerEtAl1999; @klomek2007; @klomek2019], with sexual and non-sexual assault, in particular, greatly increasing one's risk of depression [@KendlerEtAl1995; @KendlerEtAl1999]. This holds true even in a small-scale, non-Western society: among the Tsimane, Amazonian horticulturalists, depression is also associated with conflict, especially conflict involving non-kin [@StieglitzEtAl2015a]. See @SymeHagen2021 for a review of the association of depression with anger and conflict.

Other notable depression risk factors, like loss of a loved one or severe or prolonged illness, might seem less related to conflict. In these situations, however, the fitness costs that stem from reduced access to resources could be mitigated with help from social partners [@SugiyamaSugiyama2003]. However, social partners might not be able, or want, to provide more investment than they already are. Therefore, problems whose solutions require substantial investment or other changes on the part of social partners will often involve social conflict even if they did not start that way [@Hagen2003]. Indeed, there is increasing evidence that loss of a loved one is often followed by increased family conflict [see @SymeHagen2021 for a brief review].

When need is private information and there are conflicts with social partners, “cheap” signals of need, such as sad expressions and crying, may often not be believed when providing support is costly. We argue next that in this common situation, some of the most harmful and mysterious symptoms of depression -- profound loss of interest in virtually all activities and suicidality -- serve as *credible* and adaptive signals of need.

@@

With a cooperative species like our own, ubiquitous conflicts of interests means that there will always be disagreement over the levels of investment in a cooperative endeavor and the division of the resulting benefits, even among closely related individuals. According to partner choice models, individuals who are dissatisfied with the terms of cooperation can switch partners [@hammerstein2016], e.g., workers unhappy with their pay can look for a better job. In many cases, however, it is difficult or impossible to switch partners. Spouses who are dissatisfied with their partner's investment in their new infant, for example, cannot easily find a different partner to invest in that infant [@Hagen:1999]. Similarly, an adolescent who is dissatisfied with her parent's investment in her cannot easily find other parents who were willing to invest more, nor could parents easily produce another adolescent. In these latter examples, and many cooperative endeavors central to human biological fitness, all parties have monopoly power over the benefits they bring to the endeavor -- no one is easily replaced [@Hagen2002; @Hagen2003]. Such interdependence is increasingly recognized to be important to the evolution of cooperation in humans and other animals [@roberts2005; @tomasello2012; @aktipis2018; @balliet2017].

@Hagen2003 proposed that physical aggression and core depression symptoms like loss of interest in virtually all activities were complementary strategies to resolve conflicts in interdependent relationships. @SellEtAl2009 found that physically formidable individuals were more prone to anger, prevailed more in conflicts of interest, and considered themselves entitled to better treatment. Physically or socially weaker individuals, though, are not without options to resolve conflicts in their favor. An individual with monopoly power over the benefits she contributes to a critical cooperative endeavor can withhold those benefits, or put them at risk, until her partners change their behaviors in ways that benefit her. As depression often involves a profound loss of interest in virtually all activities that can jeopardize one's productivity [@dsm5], it might therefore be an evolved bargaining strategy for relatively powerless individuals in the wake of adversity and social conflict [@Hagen2003; @SymeHagen2021; see also @Watson:2002].

Bargaining models assume that delaying cooperation is costly so that there is an incentive to quickly agree on a division of benefits, especially for those who highly value the fruits of the cooperative endeavor. In a classic non-cooperative game theory model of bargaining, @rubinstein1982 showed that two parties can come to an immediate agreement over division of benefits despite conflicts of interest if the parties' valuations of cooperation are not private information: in this case, each party knows exactly what division of benefits the other will accept, and can therefore make that offer immediately, avoiding the cost of delay.

If valuations are private information, however, costly delays might be unavoidable because each party has an incentive to request more than their actual valuation, and to reject the likely inflated requests from partners, leading to multiple rounds of bargaining. Models of bargaining with private information have a close relationship to models of credible signaling. A willingness to delay (i.e., refuse offers) credibly reveals one's low valuation of the endeavor, and therefore genuine need -- the benefit of waiting for a better offer outweighs the low cost of delay. Eagerness to reach a deal, on the other hand, credibly reveals a high valuation -- the benefit of waiting for a better offer does not outweigh the higher cost of delay. Once valuations are known, the game reduces to the one in @rubinstein1982, and the parties can reach an agreement on the division of benefits [@KennanWilson1993]. Delays also typically require that additional factors come into play [@feinberg2005 and references therein].

@@

Theoretical models of depression must account for suicidality.Suicidal ideation is one of nine diagnostic criteria for a major depressive episode (MDE) [@dsm5] and is associated with depression across cultures [@haroz2017]; depression is a major risk factor for suicidal behavior [@hawton2013]; and suicidality has been invoked to justify grief-related depression symptoms as psychopathological [e.g., @pies2014].

Anthropology, in contrast, has long viewed suicidality as largely the result of social problems. Early in the field's history, anthropologists reported on suicide attempts and deaths in the small-scale societies that serve as models for the types of societies in which humans evolved. Suicide, they found, was commonly a form of protest, revenge, and/or appeal [@firth1936; @Firth:1961; @Malinowski:1932; @niehaus2012]. Some ethnographers emphasized suicide as a form of anger or social pressure [@Hezel:1987; @Giddens:1964], whereas others emphasized the powerlessness of suicide victims [@Counts:1980].

@@

As with depression, there are negative social responses to suicidal behavior. Most studies examining responses to those who have survived suicide attempts have focused on stigmatization. In these studies, perceptions of survivors as being weak, selfish, mentally ill, and antisocial are commonly reported [@BatterhamEtAl2013; @TzengLipson2004], with stigmatization often being found within and outside one's social network [@FreyEtAl2016; @ScoccoEtAl2012].

@@

We argue that among those who lack better options, aversive depression symptoms that put one's value to others at risk, such as loss of interest and suicidality, credibly signal low valuation of the current efforts of social partners and motivate them to provide more support so as to end the aversive depressive behaviors [for similar views, see @andrews2006; @HagenEtAl2008; @nock2008; @Stengel1956; @Rosenthal1993; @firth1936; @Firth:1961; @farberow1961].

@@
